### 2011-11-06 doc/pldi2012/highlevel.tex
Some edits to the execution model and high level runtime

\section{High-Level Runtime} 
\label{sec:highlevel}
The high-level runtime is responsible for guaranteeing the semantics 
of programs written in the Legion programming model while at the same
time extracting high performance.  This goal is challenging because
the high-level runtime must be capable of operating on top of many 
different architectures abstracted by the low-level runtime.  This 
includes those with very large latencies for transferring data such
as distributed memory clusters.  To hide these latencies, the high
level runtime uses a deffered model of execution made possible by the
event model of the low-level runtime.  By deffering execution, the 
runtime can hide the latencies of data and task movement and acieve
good throughput and high performance.

However, a deffered execution model presents challenges to implementing
the semantics of the Legion programming model.  Deffered execution
does not necessarily imply that tasks will be executed in the order in
which they are scheduled, but only in the order in which low-level
event dependencies are created.  To orchestrate the execution of Legion
programs in this environment the high-level runtime is architected
in a manner similar to an out-of-order hardware processor.  There are
multiple stages to executing a task which allows the runtime to extract
as much task-level parallelism as possible from an application whithin
the constraints of the Legion programming model.  We now investigate
each of these stages in further detail.

\subsection{Dependence Analysis} 
\label{subsec:depanalysis}
The first step in the execution of a task is dependence analysis.  When
a task is registered with the high level runtime the task call includes
information about both the logical regions that the task will require
when it is executed as well as the read-write access and coherence properties
for each of the regions.  The runtime uses this information to perform
dependence detection with each of the previous tasks that have been registered
in the same parent task.  Dependence detection only needs to be performed at
the scope of an enclosing task because the semantics of the Legion programming
model ensure that all tasks can only use regions which are subregions of
regions that the parent task uses.  This implies that if a child task were
to depend on a task at a higher level of the task tree, then the
parent task would also have had a dependence.

Lemma: If a task {\tt t} with ancestor task {\tt p} depends on task {\tt t'}
a sibling task of {\tt p}, then task {\tt p} must depend on task {\tt t'}.

This property of the Legion programming model enables the runtime to only 
have to perform dependence analysis between tasks which share the same 
parent task.  By restricting dependence analysis to tasks which share the same
parent task, the runtime can perform dependence analysis locally and not
have to be concerned with tasks being created in other parts of the machine.

When performing dependence analysis the runtime relies on its knowledge about
the relationships between logical regions and partitions.  We first describe
the the components to dependence analysis in sections \ref{subsec:regiontree},
\ref{subsec:cohdep}, and \ref{absinsts}.  We then describe the algorithm for performing 
dependence analysis on this data structures in section \ref{subsec:depdetect}.

\subsubsection{Region Trees}
\label{subsec:regiontree}
To detect dependences between the tasks, the runtime leverages its knowledge
about logical regions and their partitions.  To create a new logical region
or partition, the application must invoke the corresponding call in the
runtime.  As these calls occur, the runtime constructs a data strcture
called a {\em region tree}.  A region tree describes the relationship between
logical regions and partitions.  A region tree contains two types of nodes:
region nodes and partition nodes.  Every region tree is rooted with a region
node and alternates between region and partition nodes at each corresponding level.
A region node represents a specific logical region and tracks the set of
partitions of that logical region.  Similarly a partition node represents 
a specific partitioning of a logical region and keeps track of the logical regions 
which are subregions of the partition.

\subsubsection{Dependencies From Coherence Properties}
\label{subsec:cohdep}
To detect dependencies between any two tasks using a logical region, we leverage
the information contained in the access and cohrence properties provided by the programmer
for how the region is to be used.  Dependency checks are performed pair-wise between regions
to determine whether one task depends on the other.  When checking for
dependencies we first examine the access property.  If both tasks are reads, then 
there is never a dependence.  If at least one of them is a write we use table
\ref{tab:depsdetect} to determine whether there is a dependence.

\begin{center}
\begin{table}
\begin{tabular}{|c|c|c|c|c|} \hline
             & Exclusive & Atomic   & Simultaneous & Relaxed \\ \hline 
Exclusive    & Dependence & Dependence & Dependence & Dependence \\ \hline
Atomic       & Dependence & Same Instance & Contingent & Contingent \\ \hline
Simultaneous & Dependence & Contingent & Same Instance & None \\ \hline
Relaxed      & Dependence & Contingent & None & None \\ \hline
\end{tabular}
\caption{Dependence detection cases when at least one access is a write.\label{tab:depsdetect}}
\end{table}
\end{center}
In this table, {\em Dependence} indicates that there will always be a dependence between
regions while {\em None} indicates that there will never be a dependence regardless
of mapping decisions.  In the case of {\em Same Instance} there is a dependence contingent
upon a mapping decision.  If both tasks map to the same physical instance (described in 
more detail in section \ref{subsec:instmang}) then there is no dependence, which if
they map to different instances there is a dependence as the system will have to generate
a copy between the two tasks running.

For the case of {\tt Contingent} there are two special cases that we handle.
One special case is Write-After-Read (WAR) dependencies.  For the WAR case, if neither coherence property
is {\em Exclusive} then we can ensure that different physical instances
are used to avoid an anti-dependence and increase parallelism.  If either property is {\em Exclusive} then the semantics
of the programming model mandate that we detect a dependence and serialize the task.

@@

\subsubsection{Abstract Instances}
\label{subsec:absinsts}
For the cases where tasks do not depend, they are allowed to create different physical
instances of the same logical region.  The decision to do this however does not
happen until the mapping phase of a task execution.  However, we need to have a place
holder that is capable of representing all the possible valid physical instances
from which to choose even if those instances do not exist because the prior
tasks have not been mapped.  We call these place holders {\tt abstract instances}.

Abstract instances represent the set of possible valid physical instances which a
task will be capable of selecting from or adding to when it is mapped.  The value
of abstract instances is that they allow us to remember which physical instances
are valuable while continuing to perform dependence detection on newly registered
tasks.  In the analogy to out-of-order processors abstract instances allow us
to perform the same operation as register-renaming of instructions to avoid
anti-dependences and to remember actual dependences even though our dependence detection
analysis may run ahead of the actual execution of tasks.

\subsubsection{Dependence Detection}
\label{subsec:depdetect}
When a task is registered, the runtime
will perform dependence detection for every region that the task requests with
prior tasks that have been registered in the same parent task context.  Since
tasks are registered in program order we will always perform dependence detection
in program order which guarantees correctness even if tasks are executed out-of-order.  
Performing dependence detection in program order is also annalogous to how an out of order
processor works.  Out-of-order processors load instructions in order and only after
analyzing their instructions and register dependencies do they execute instructions
out of order.  The regions that a task requires are annalogous
to the registers listed in an instruction and the goal of the high level
runtime is to detect the same dependences and use them to enforce scheudling constraints
upon a task.  However, there is a catch to this analogy: in the Legion programming
model two different logical regions are not necessarily independent from each other.  In
order to do the dependence detection we rely on the forest of region tree data structures to be
able to detect dependences between tasks which require different logical regions.

To perform dependence detection, we first find the {\em root region} for the 
requested region.  The root region is the ancestor region for which the parent task
has privliges.  Note that the semantics of the Legion programming model state that
we must always be able to find such a root region or else an error is thrown.

Once we have found the root region for the requested region, we then proceed to walk
down the region tree from the node corresponding to the root region to the node
for the requested region.  At each node, we determine if any potential dependences exist.
Determining whether dependencies exist involves checking information both about the current
node being visited as well as whether there are any potential dependences farther down
in the region tree.  We now describe the decision procedure performed at both region
nodes and partition nodes in the region trees.

@@

\subsubsection{Implicit Dataflow Construction}
\label{subsec:dataflow}
In addition to computing the dependences between tasks, we also have to compute where
a task is permitted to acquire its data from when it goes to execute.
It is insufficient to simply record the producer task as there
may be multiple active tasks for a logical region, all of which may produce different physical
instances.  In addition, it is likely that the producer tasks have not been mapped
yet which makes it impossible to know the set of physical instances from which to pull data.

>>>>

\section{High-Level Runtime} 
\label{sec:highlevel}
In this section we describe a high level runtime system
that is designed to implement the Legion programming
and execution models while remaining machine agnostic.  
The high level runtime system operates on top of the 
low-level runtime which hides machine specific details
and is described in detail in section \ref{sec:lowlevel}.

The biggest challenge to a implementing the high level
runtime is to avoid the latencies associated with distributed
memory systems and deep memory hierarchies.  To aid
in this battle, the high level runtime employs two useful 
scheduling techniques.  The first useful technique was proved
in section \ref{sec:exec}: all scheduling decisions for a
task can be made locally in the context in which the
task was created.  The second technique is a deferred execution
model that allows the runtime to hide the latencies associated
with moving data through deep memory hierarchies.

%The high-level runtime is responsible for guaranteeing the semantics 
%of programs written in the Legion programming model while at the same
%time extracting high performance.  This goal is challenging because
%the high-level runtime must be capable of operating on top of many 
%different architectures abstracted by the low-level runtime.  This 
%includes those with very large latencies for transferring data such
%as distributed memory clusters.  To hide these latencies, the high
%level runtime uses a deffered model of execution made possible by the
%event model of the low-level runtime.  By deffering execution, the 
%runtime can hide the latencies of data and task movement and acieve
%good throughput and high performance.

A deffered execution model however presents challenges to implementing
the semantics of the Legion programming model.  Deffered execution
does not necessarily imply that tasks will be executed in the order in
which they are scheduled, but only in the order in which dependences
are expressed.  To orchestrate the execution of Legion
programs in this environment the high-level runtime is architected
in a manner similar to an out-of-order hardware processor, but at
a coarser granularity with task corresponding to instructions and
regions corresponding to registers.  There are multiple stages to 
executing a task which allows the runtime to extract
as much task-level parallelism as possible from an application whithin
the constraints of the Legion programming model.  To illustrate
the implementation of the high level runtime we now describe the
details of how each stage of task execution is performed.

\subsection{Dependence Analysis} 
\label{subsec:depanalysis}
The first step in the execution of a task is dependence analysis.  When
a task is registered with the high level runtime the task call includes
information about both the logical regions that the task will require
when it is executed as well as the read-write permissions and coherence properties
for each of the regions.  The runtime uses this information to perform
dependence detection with each of the previous tasks that have been registered
in the same context.  Dependence detection only needs to be performed at
the scope of an enclosing context as was proved in section \ref{sec:exec}.

We initially describe how dependences are detected from logical regions
in section \ref{subsec:cohdep}.  We then describe the algorithm that
is used to perform dependence detection in section \ref{subsec:depdetect}.
Section \ref{subsec:dataflow} describes how data is versioned correctly
during deffered execution.  For concreteness we describe how dependence
analysis is performed for the circuit simulation from section \ref{sec:ex}.

%When performing dependence analysis the runtime relies on its knowledge about
%the relationships between logical regions and partitions.  We first describe
%the the components to dependence analysis in sections \ref{subsec:regiontree},
%\ref{subsec:cohdep}, and \ref{absinsts}.  We then describe the algorithm for performing 
%dependence analysis on this data structures in section \ref{subsec:depdetect}.

%\subsubsection{Region Trees}
%\label{subsec:regiontree}
%To detect dependences between the tasks, the runtime leverages its knowledge
%about logical regions and their partitions.  To create a new logical region
%or partition, the application must invoke the corresponding call in the
%runtime.  As these calls occur, the runtime constructs a data strcture
%called a {\em region tree}.  A region tree describes the relationship between
%logical regions and partitions.  A region tree contains two types of nodes:
%region nodes and partition nodes.  Every region tree is rooted with a region
%node and alternates between region and partition nodes at each corresponding level.
%A region node represents a specific logical region and tracks the set of
%partitions of that logical region.  Similarly a partition node represents 
%a specific partitioning of a logical region and keeps track of the logical regions 
%which are subregions of the partition.

\subsubsection{Dependencies From Coherence Properties}
\label{subsec:cohdep}
To detect dependencies between any two tasks using a logical region, we leverage
the information contained in the permission and cohrence properties provided by the
task call.  Dependency checks are performed pair-wise between regions
to determine whether one task depends on the other.  When checking for
dependencies we first examine the permission.  If both tasks are reads, then 
there is never a dependence.  If at least one of them is a write we use table
\ref{tab:depsdetect} to determine whether there is a dependence.
\begin{center}
\begin{table}
\begin{tabular}{|c|c|c|c|c|} \hline
             & Exclusive & Atomic   & Simultaneous & Relaxed \\ \hline 
Exclusive    & Dep & Dep & Dep & Dep \\ \hline
Atomic       & Dep & Same & Cont & Cont \\ \hline
Simultaneous & Dep & Cont & Same & None \\ \hline
Relaxed      & Dep & Cont & None & None \\ \hline
\end{tabular}
\caption{Dependence detection cases when at least one access is a write. {\tt Dep}
indicates a dependence is detected.  {\tt Same} indicates a dependence is detected
only if the task uses different instances.  {\tt Cont} means that the dependence
is contingent upon the permissions.  {\tt None} means there will never
be a conflict.\label{tab:depsdetect}}
\end{table}
\end{center}
%In this table, {\em Dependence} indicates that there will always be a dependence between
%regions while {\em None} indicates that there will never be a dependence regardless
%of mapping decisions.  In the case of {\em Same Instance} there is a dependence contingent
%upon a mapping decision.  If both tasks map to the same physical instance (described in 
%more detail in section \ref{subsec:instmang}) then there is no dependence, which if
%they map to different instances there is a dependence as the system will have to generate
%a copy between the two tasks running.
For the case of {\tt Contingent} there are two special cases that we handle.
One special case is Write-After-Read (WAR) dependencies.  For the WAR case, if neither coherence property
is {\em Exclusive} then we can ensure that different physical instances
are used to avoid an anti-dependence and increase parallelism.  If either property is {\em Exclusive} then the semantics
of the programming model mandate that we detect a dependence and serialize the task.

@@

%\subsubsection{Abstract Instances}
%\label{subsec:absinsts}
%For the cases where tasks do not depend, they are allowed to create different physical
%instances of the same logical region.  The decision to do this however does not
%happen until the mapping phase of a task execution.  However, we need to have a place
%holder that is capable of representing all the possible valid physical instances
%from which to choose even if those instances do not exist because the prior
%tasks have not been mapped.  We call these place holders {\tt abstract instances}.

%Abstract instances represent the set of possible valid physical instances which a
%task will be capable of selecting from or adding to when it is mapped.  The value
%of abstract instances is that they allow us to remember which physical instances
%are valuable while continuing to perform dependence detection on newly registered
%tasks.  In the analogy to out-of-order processors abstract instances allow us
%to perform the same operation as register-renaming of instructions to avoid
%anti-dependences and to remember actual dependences even though our dependence detection
%analysis may run ahead of the actual execution of tasks.

\subsubsection{Dependence Detection}
\label{subsec:depdetect}
When a task is registered, the runtime
will perform dependence detection for every region that the task requests with
prior tasks that have been registered in the same context.  Since
tasks are registered in program order we will always perform dependence detection
in program order which guarantees correctness even if tasks are executed out-of-order.  
Performing dependence detection in program order is annalogous to how an out of order
processor works.  Out-of-order processors load instructions in order and only after
analyzing their instructions and register dependencies do they execute instructions
out of order.  The regions that a task requires are annalogous
to the registers listed in an instruction and the goal of the high level
runtime is to detect the same dependences and use them to enforce scheudling constraints
upon a task.  However, there is a catch to this analogy: in the Legion programming
model two different logical regions are not necessarily independent from each other.  In
order to do the dependence detection we rely on the forest of regions to be
able to detect dependences between tasks which require different logical regions.

To perform dependence detection, we first find the {\em root region} for the 
requested region.  The root region is the ancestor region for which the parent task
has privliges.  Once we have found the root region for the requested region, we then proceed to walk
down the region tree from the root region to the node
for the requested region.  At each node, we determine if any potential dependences exist.
Determining whether dependencies exist involves checking information both about the current
node being visited as well as whether there are any potential dependences farther down
in the region tree.  We now describe the decision procedure performed at both region
nodes and partition nodes in the region trees.

@@

\subsubsection{Implicit Dataflow Construction}
\label{subsec:dataflow}
In addition to computing the dependences between tasks, we also have to compute the version
of data that a task is permitted to use when it goes to execute.
It is insufficient to simply record the producer task as there
may be multiple active tasks for a logical region, all of which may produce different physical
instances.  In addition, it is likely that the producer tasks have not been mapped
yet which makes it impossible to know the set of physical instances from which to pull data.
